Computing likelihoods under the SMC for a general class of ARGs.

### 2023-12-12 paper.tex
update intro

\section{Abstract}
This means there currently is no way to connect the output of the diverse 
range of ARG inference methods back 
to a model that would allow us to perform a probabilistic exploration of the 
ARG space around the point estimate, or likelihood-based inference

@@

\subsection{ARGs are awesome, or general intro about approximations}
% approximation: level of detail
%Importance of ARGs: ARGs as object of theoretical reasoning vs what can be known\\
%difference between precision/detail and accuracy. High precision can 
%give a false sense of certainty in the provided estimate].

In population genetics, the questions we try to answer are generally 
about the unobserved processes that gave rise to the genetic variation 
we can observe given a set of samples. 
The coalescent with recombination 
Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
the generative model that allows us to understand the genealogy 
underlying a particular sampleset. 
Datastructure that captures it: ARG.
For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history  
where all events are uniquely positioned both in time
and along the genome. [However see Wong et al. 2023]. 
% say something about the difficulty of precise ARG inference (see more scalable ARG inference later on).

Although ARG, ...,
something about the level of detail.
% easy by which the CwR can be used to simulate is not matched by the usability of the model for inference

@@

Instead, approximations of the coalescence with recombination (CwR) have  
driven our ability to perform statistical genomic inference. Two approximations 
in particular have proven to be instrumental [...]. Tightly linked to their 
ability to reformulate these inference problems as hidden markov Models (HMM) 
and use the well-known efficient HMM-algorithms. 
Both: 1/ Left to right, and 2/ discetization 
of state space by emumerating tree topologies and/or discretising time.\\

To improve the ability to simulate the CwR, McVean and Cardin 
provided an approximation of the CwR that is both Markovian backwards in time 
as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
By sacrificing precision on our ability to capture long range linkage 
information, the sequentially Markovian coalescent (SMC) 
became the engine of many powerful inference methods [PSMC]. 
A later improvement, the SMC', closed the gap between the SMC and the 
CwR even further.

A second key approximation is not a generative model, instead the 
copying model \citep{li_2003_modeling} was specifically designed to 
generate CwR-like patterns. 
Rather than providing a mathematical description of the genealogies 
underlying the observed data it pragmatically describes the relationship 
of a single genome to a larger set of samples as the result of an 
incomplete copying process. This heuristic approach has been successfully 
applied for demographic inference [and other things: ADD] 
\citep{sheehan_2013_estimating, steinrucken_2018}.

Both approaches have further formed the basis of the recent advancements in 
scalable ARG inference methods [references]. 
First real breakthrough, ARGweaver, samples an ARG from the posterior distribution 
given the observed variational data. Although require limiting the set of 
possible node times to be compatible with HMM algorithms, the ARGs returned 
by ARGweaver are fully precise and detailed in the sense that all 
recombination events are explictly inferred. 
This is where the real big difference lies with all other scalable methods. 
Although all different, all these methods share the fact that the ARGs they infer 
are approximate structures. The ARGs they infer have varying degrees of detail

\subsection{Recent progress in ARGs through approximate structures}

% bring nuance in the distinction between detail and approximation

% heuristic methods guided by coalescent principles (Li and Stephens)

@@

% what is an approximate structure: implicit recombination, polytomies or not, ...
They lack the full detail: implicit recombination, ...
In particular, recombination is dealt with implicitly. Recombination events are 
not directly inferred and thus not dated. Tree topology changes from one local 
tree to the next without explicit inference of the one or more recombination 
events required to explain the inferred tree transition.

Note that approximate structure need not be an issue, the fully precise and detailed 
ARG can never be known. Instead, the main question is how to connect these 
approximate structures to the existing generative models.

% Existing formalisms assume that an ARG is known in full detail
These recent developments have also meant however that
we are now using the term ARG for a varying range of approximate structures 
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).

@@

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of

$N_e$-changes as well as multiple populations and migration can be readily incorporated 
in the current algorithm.\\

@@

% 4/ quantifying uncertainty [unfinished]:
Both of these future directions tie in with the desired ability of any ARG-related method to 
quantify the uncertainty associated with

>>>>

\section{Abstract}
There currently is no way to connect the output of the diverse 
range of ARG inference methods back 
to a model that would allow us to perform a probabilistic exploration of the 
ARG space around the point estimate, or likelihood-based inference.

@@

\subsection{ARGs are awesome, or general intro about approximations}

[TL;DR] Importance of ARGs.
There is a discrepancy between ARGs as object of theoretical reasoning 
vs what can be known. Explain the difference between precision/detail and accuracy. 
High precision can give a false sense of certainty in the provided estimate.

In population genetics, the questions we try to answer are generally 
about the unobserved processes that gave rise to the genetic variation 
we can observe given a set of samples. 
The coalescent with recombination 
Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
the generative model that allows us to understand the genealogy 
underlying a particular sampleset. 
Datastructure that captures it: ARG.

For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history where all events are uniquely positioned 
both in time and along the genome. Yet the ease by which the CwR can be simulated,
and therefore the ease by which an ARG can be generated, is not not matched by 
the usability of the model for inference.

@@

Instead, approximations of the coalescence with recombination (CwR) have  
driven our ability to perform statistical genomic inference. Two approximations 
in particular have proven to be instrumental: the sequentially Markovian coalescent, 
and its extensions (SMC) on the one hand, and the Li and Stephens model on the other. 
The impact of both models is tightly linked to the fact that they have enabled 
the reformulation of inference problems as hidden Markov models (HMM) and thus 
the use of the well-known efficient HMM-algorithms.
%Both: 1/ Left to right, and 2/ discetization 
%of state space by emumerating tree topologies and/or discretising time.\\

To improve the ability to simulate the CwR, McVean and Cardin 
provided an approximation of the CwR that is both Markovian backwards in time 
as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
By sacrificing precision on our ability to capture long range linkage 
information, the SMC became the engine of many powerful inference methods [PSMC]. 
A later improvement, the SMC', closed the gap between the SMC and the 
CwR even further.

The second key approximation is not a generative model. Rather than providing 
a mathematical description of the genealogies underlying the observed data 
the Li and Stephens or copying model pragmatically details the relationship 
of a single genome to a larger set of genomes as the result of an 
incomplete copying process \citep{li_2003_modeling}. This heuristic approach 
has been successfully applied for demographic inference [and other things: ADD] 
\citep{sheehan_2013_estimating, steinrucken_2018}.

Both approaches have further formed the basis (of many) of the recent advancements in 
scalable ARG inference methods [references]. A real distinction can be made however 
between \argweaver, the first real breakthrough in scalable ARG inference, and the 
more recent methods. The ARGs sampled by \argweaver from the posterior distribution 
given the observed variational data, are all fully precise and detailed in the sense 
that all events are explicitly inferred. Note however that the set of possible node times 
has to be limited to enable the usage of an HMM. This fully detailed inference holds 
for recombination, but also for migration events \cite{hubisz_2020_mapping}. 
All other methods, on the other hand, derive their scalability at least to a 
certain degree from the fact that 
the ARGs they infer are approximate structures. In particular, recombination is an 
emergent property of the inferred ARGs explaining tree topology changes going 
from one local tree to the next. The recombination events themselves are not 
explicitly inferred and are thus not part of the resulting data structure.

\subsection{Recent progress in ARGs through approximate structures}
% onto other features

The computational gains from not having to pinpoint recombination events tie in 
with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022].

Always events that cannot be known. Different ways of dealing with this.
Some approaches have embraced the approximate nature of the resulting ARG to for 
example highlight the inability to resolve certain splits (polytomies).

others rigorous treatment of uncertainty (Arbores, ARGinfer): fully precise about recombination.

% onto ARG cannot be fully known
Note that approximate structure need not be an issue, the fully precise and detailed 
ARG can never be known. Instead, the main question is how to connect these 
approximate structures to existing models.
While this is not a problem when describing the results of simulations 
(where all details are perfectly known), it is a major issue when we wish to formally 
describe the output of inference methods, particularly as datasets approach the population 
scale. 
As discussed in the introduction, the precise details of all 
events in these vast ARGs can never be known, and a data structure that enforces 
complete precision is therefore an impediment to progress.

@@

These recent developments have also meant however that
we are now using the term ARG for a varying range of approximate structures 
as generated by these different inference methods \citep{wong_a-general_2023}.

Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).
% Existing formalisms assume that an ARG is known in full detail

@@

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of [...] 
$N_e$-changes as well as multiple populations and migration can be readily incorporated 
in the current algorithm.\\

@@

% 4/ quantifying uncertainty [unfinished]:
Both of these future directions tie in with the desired ability of any ARG-related method to 
quantify the uncertainty associated with ARG inference
